Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-01-09 Manuscript/Figures/figure_legends.Rmd
edits

**Figure 1. Ensemble Framework.** A diagram of the cell line specific ensemble framework detailing the various combination of inputs used in the model building pipeline. The minoirty and majority classes were represented by the genomic bins that overlapped with flanked TAD boundaries or did not, respectively. Functional genomic annotations obtained from the UCSC Genome Browser Database composed the feature space. The data was split into a 7:3 training to testing set ratio. A feature engineering step was included to isolate the spatial relationship, of increasing granularity, between annotations and genomic bins. Data re-sampling and regularization was applied to the training data. Lastly, a random forest classification algorithm was implemeted and validated on the testing data.

<!--
A given system within the ensemble framework included the total number of genomic bins, labeled as having been overlapped with a flanked TAD boundary ($Y=1$) or not ($Y=0$). The data matrix that composed the feature space was composed of predictor-type specific annotations represented by $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations. The data was then split into a 7:3 training to testing set ratio. A re-sampling technique was applied to the training set, followed by an elastic-net regularization to reduce the feature space. Lastly, a random forest classification model was performed on the re-sampled and reduced training set and validated on the held out testing set common to all ensemble systems in the framework.
-->

**Figure 2. Model Construction.** An illustration of the proposed model construction given cell line specific Hi-C data. Above is an example of a subsection of the contact matrix for chromosome 14 (at megabase scale) on GM12878 [@durand2016juicer]. The contact matrix is a visualization of the interconnectivity of the linear genome at 100 kb resolution. The boundary coordinates of TADs were identified using the Arrowhead algorithm. To account for the uncertainty in TAD boundary detection, each boundary point was flanked on both sides by resolution-specific genomic regions (bins). The response vector $Y$, used for classification, was determined by whether a genomic bin overlapped with a flanked boundary region ($Y=1$) or not ($Y=0$). The spatial relationship between positional coordinates of various functional genomic annotations, obtained from the UCSC Genome Browser Database, together with the genomic bins were then used to define the feature space of the models represented by the data matrix $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations.

**Figure 3. Predictor types.** Three separate types of predictors were considered when quantifying the association between TAD boundaries and genomic annotations. Each predictor type was used to build the feature space of specific models in the ensemble framework. (A) _Overlap Counts_ (OC) consider the total number of genomic features that overlapped with each genomic bin. (B) _Overlap Percent_ (OP) was calculated by dividing the sum of all feature widths within each specific bin by the total bin width. (C) _Distance_ ($log_{2}$ transformed) was calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest region defining a genomic annotation.

**Figure 4. SMOTE re-sampled models built on distance-type predictors yielded optimal MCCs across resolutions for GM12878.** Performances of random forest classification models built on each of the 3 predictor types: (A) OC, (B) OP, and (C) Distance. Within each predictor type, and across resolution, training data was re-sampled using different techniques including: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Figure 5. The proportional importance of optimally built predictive models was outweighed by transcription factor binding sites across resolutions for GM12878.** Stacked barplots illustrate the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors), for each resolution on GM12878. Proportions are grouped by the class of cell line specific annotations in the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).

**Figure 6. Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for GM12878** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.

**Figure 7. Top 20 most predictive transcription factor binding sites sorted by 10 kb for GM12878.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.

**Figure 8. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with 10 kb flanked TAD boundaries on GM12878** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around 10 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].

>>>>

**Figure 1. Illustrating the resolution-specific data construction and feature engineering used for downstream modeling.** (A) Unique boundaries are denoted by the blue dots. The linear genome was subsequently binned into resolution-specific intervals. The response vector **Y** was defined as a genomic bin that either overlapped with a TAD boundary ($Y = 1$) or did not ($Y = 0$). (B) Three types of associations between bins and genomic annotations (green shapes) used to build the feature space of random forest classification models include _Overlap Counts_ (OC), _Overlap Percent_ (OP), and ($log_{2}$) _Distance_.

**Figure 2. A diagram of the ensemble framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the rowise (genomic bins) and colwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)). Step 2 is to split **A** into a training and testing set according to a 7:3 ratio. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to implement the random forest algorithm and perform 5-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data for unbiased model evaluation.

### 2020-01-09 Manuscript/Supplementary/supplementary_legends.Rmd
edits

**Supplementary Figure 1. Performing $log_{2}$ transformations normalized skewed distance densities across resolution for GM12878.** When measured in basepairs, distances were highly right skewed for all resolutions: (A) 10 kb, (B) 25 kb, (C) 50 kb, and (D) 100 kb. After $log_{2}$ transforming the distances, densities became much more gaussian distributed for all resolutions: (E) 10 kb, (F) 25 kb, (G) 50 kb, and (H) 100 kb. Each density curve represents a cell line specific annotation, and its relationship with all resolution specific genomic bins.

**Supplementary Figure 2. Performing $log_{2}$ transformations normalized skewed distance densities across resolution for K562.** When measured in basepairs, distances were highly right skewed for all resolutions: (A) 10 kb, (B) 25 kb, (C) 50 kb, and (D) 100 kb. After $log_{2}$ transforming the distances, densities became much more gaussian distributed for all resolutions: (E) 10 kb, (F) 25 kb, (G) 50 kb, and (H) 100 kb. Each density curve represents a cell line specific annotation, and its relationship with all resolution specific genomic bins.

**Supplementary Table 3. Data Summaries and Imbalance Ratios.** For K562, average number of genomic bins, TADs called, and the corresponding boundaries and imbalance ratios, across each of the autosomal chromosomes (omitting chromosome 9), stratified by resolution. The resolution of the Hi-C contact matrix dictated the width of each genomic bin, with lower resolutions corresponding to wider genomic bins.

**Supplementary Table 4. Inter-Resolution ICCs.** For K562, two-way intraclass correlation coefficients between resolutions for the following quantities: Number of genomic bins, number of TADs called, number of unique TAD boundaries, and imbalance ratio. The ICC was computed across 21 of the autosomal chromosomes. All quatities, except for imbalance ratio, was correlated with data resolution. Instead, imbalance ratios remained consistent and low across resolutions.

**Table 5. Jaccard Indices.** For K562, the median (range) Jaccard index illustrating the pairwise similarity between annotations within specific groups, including chromatin segmentation states, histone modifications, and transcription factor binding sites. The Jaccard Index measures the ratio of the number of intersecting base pairs between any two sets, to the number of base pairs in the union of the two sets. The number of cell line specific annotations included within each group is also provided.

**Supplementary Table 6. Full list of aggregated performances for GM12878.** Performances were stratified by resolution, resampling technique, and predictor type, and then averaged across chromosomes.

**Supplementary Table 7. Full list of aggregated performances for K562.** Performances were stratified by resolution, resampling technique, and predictor type, and then averaged across chromosomes.

**Supplementary Figure 3. SMOTE re-sampled models built on distance-type predictors yielded optimal MCCs across resolutions for K562.** Performances of random forest classification models built on each of the 3 predictor types: (A) OC, (B) OP, and (C) Distance. Within each predictor type, and across resolution, training data was re-sampled using different techniques including: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.

**Supplementary Figure 4. The proportional importance of optimally built predictive models was outweighed by transcription factor binding sites across resolutions for K562.** Stacked barplots illustrate the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors), for each resolution on GM12878. Proportions are grouped by the class of cell line specific annotations in the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).

**Supplementary Figure 5. Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for K562** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.

**Supplementary Figure 6. Top 20 most predictive transcription factor binding sites sorted by 10 kb for K562.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.

**Supplementary Figure 9. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with varying units of flanked TAD boundaries on GM12878** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around (A) 25 kb, (B) 50 kb, and (C) 100 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].

**Supplementary Figure 10. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with varying units of flanked TAD boundaries on K562** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around (A) 10 kb, (B) 25 kb, (C) 50 kb, and (D) 100 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, and ZNF143. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].

<!--
**Supplementary Figure 12. Comparing $log_{2}$ distance profiles of top predictive transcription factor binding sites by bin class for 25 kb resolution, on GM12878.** Violin plots (top) and plots of differences in quantiles (bottom) between genomic bin class for 4 of the top predictive TFBS determined by the best performing model, including: SMC3 (A), ZNF143 (B), CTCF (C), and RAD21 (D). For the plots of differences in quantiles, the reference group are the genomic bins that did not overlap with a flanked TAD boundary. That is, quantile differences above 0 indicate that bins that overlapped with flanked TAD boundaries exhibited smaller quantiles.
-->

>>>>

**Supplementary Table 3.** Domain data and class imbalance summaries across resolutions for GM12878.

**Supplementary Figure 1. Performing $log_{2}$ transformations normalized skewed genomic distance densities across all resolutions.** When measured in basepairs, distances were highly right skewed for all resolutions for both the (A) GM12878 and (B) K562 cell lines. After $log_{2}$ transforming the distances, densities became much more gaussian distributed for all resolutions.

### 2020-01-09 Manuscript/manuscript.Rmd
edits

Genomic annotation data in the form of cell type-specific cis-regulatory elements (CSS), histone modifications (HM), and transcription factor binding sites (TFBS) were obtained from the UCSC Genome Browser Database (GRCh37/hg19 human genome assembly) [@Tyner:2017aa] (Supplementary Table 2).

@@

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. <!--After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin]),--> We obtained a total of 16133, 14824, 8081, 3560, and 919 unique TAD boundaries for the GM12878 cell line, across each of the five resolutions respectively (Table 1). Similar numbers of TAD boundaries were found for the K562 cell line. To account for the variability in boundary location due to varying resolution, we developed a resolution-specific approach toward constructing TAD versus non-TAD boundary regions on the linear genome (See Methods), labeled as a response vector **Y** (Figure 1A). TAD boundary regions were defined as regions on the linear genome where a genomic bin overlapped with a flanked TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as the regions that did not overlap with a flanked TAD boundary ($Y = 0$). On average, we found that only $18.0 \pm 4.6 \%$ of TAD boundary regions overlapped between both cell lines. Given the cell type specificity of TAD boundaries, we opted to build cell line specific models for predicting TAD boundary regions. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. In summary, we developed a framework enabling resolution-specific prediction of TAD boundaries, with resolution-specific bins defining the response variable.

| Resolution | Total number of called TADs | Total number of unique TAD boundaries | Total number of Genomic Bins | Class Imbalance |
|:----------:|:---------------------------:|:-------------------------------------:|:----------------------------:|:---------------:|
|5 kb        |8052                         | 15468                                 | 535363                       | 0.03            |
|10 kb	     |7676 	                       | 14253                                 | 267682                       | 0.05            |
|25 kb	     |4670                         | 8363                                  | 107073                       | 0.08            |
|50 kb	     |2349	                       | 4224                                  | 53537                        | 0.08            |
|100 kb	     |1031                         | 1883                                  | 26768                        | 0.07            |
Table: Domain data and class imbalance summaries across resolutions for GM12878.

@@

A total of 78 cell type specific functional genomic annotations for the GM12878 and K562 cell line were used to build the feature space of downstream predictive models (Supplementary Table 1). The annotations could be broken down into three sets including histone modifications (HM), chromatin state segmentations (ChromHMM), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1B). Counts of overlaps (**overlap count**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**overlap percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of distances (Supplementary Figure 1). These feature engineering steps were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

@@

Given that the number of bins annotated as TAD boundaries (TAD boundary regions) is much smaller than the number of non-TAD boundary bins, we investigated the class imbalance (CI) featured at each resolution. The CI was defined as the ratio of TAD boundary regions ($Y=1$) versus the sum of both TAD and non-TAD boundary regions and ranging from 0 to 0.50. The CI was found to be invariant to resolution, remaining consistently low across chromosomes, with an averages of 0.06, 0.10, 0.14, 0.12, and 0.06 across each resolutions for GM12878 (Table 1). Similar class imbalances were found for the K562 cell line. To account for such imbalance, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. **Random over-sampling** (ROS) was defined as sampling with replacement from TAD boundary regions. **Random under-sampling** (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered **Synthetic minority over-sampling technique** (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

## Ensemble framework for TAD boundary prediction

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using one of the predictor type *P = {OC, OP, Distance}*, and the re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 10-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 = 2520$) models. <!--Additional performance metrics are reported in the Supplementary materials.-->

<!--
![**A diagram of the ensemble framework detailing the combination of inputs used in the model building pipeline.** Each cell line (GM12878, K562) specific model was composed of a combination of different data level characteristics including resolution (5 kb, 10 kb, 25 kb, 50 kb, 100 kb), predictor type (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)), and re-sampling technique (no re-sampling (None), random over-sampling (ROS), random under-sampling (RUS), synthetic minority over-sampling technique (SMOTE)). The three types of genomic annotations, histone modification (HM), chromating state segmentations (ChromHMM), transcription factor binding sites (TFBS), were used as predictors.](Figures/figure3_ensemble_framework_V3.png)
-->

>>>>

Genomic annotation data in the form of cell type-specific cis-regulatory elements (CRE), histone modifications (HM), and transcription factor binding sites (TFBS) were obtained from the UCSC Genome Browser Database (GRCh37/hg19 human genome assembly) [@Tyner:2017aa] (Supplementary Table 2).

@@

We analyzed publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. <!--After removing TADs that exceeded 2 md in length (biologically plausible maximum TAD size [@dixon2012topological; @sofueva2013cohesin]),--> The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material. We obtained a total of 8052, 7676, 4670, 2349, and 1031 called TADs across each resolution respectively (Table 1). From the called TADs we obtained a total of 15468, 14253, 8363, 4224, and 1883 unique TAD boundaries (Table 2). The resolution-specific approach we developed to create TAD vs non-TAD boundary regions amenable for binary classification (See Methods) resulted in a total of 535363, 267682, 107073, 53537, and 26768 genomic bins. TAD boundary regions were defined as regions on the linear genome where a genomic bin contained a called TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as genomic bins that did not contain a TAD boundary ($Y = 0$) (Figure 1A).

Under this construction, the class imbalance among the genomic bins was found to be both severe and invariant toward resolution. Specifically, we found the imbalance between TAD boundary and non-TAD boundary regions to be 0.03, 0.05, 0.08, 0.08, and 0.07 for each resolution respectively (Table 2). Furthermore, the data summaries presented in Table 1 were found to be similar to those exibited by the K562 cell line (Supplementary Table 3).

| Resolution | Total number of called TADs | Total number of unique TAD boundaries | Total number of genomic bins | Class imbalance |
|:----------:|:---------------------------:|:-------------------------------------:|:----------------------------:|:---------------:|
|5 kb        |8052                         | 15468                                 | 535363                       | 0.03            |
|10 kb	     |7676 	                       | 14253                                 | 267682                       | 0.05            |
|25 kb	     |4670                         | 8363                                  | 107073                       | 0.08            |
|50 kb	     |2349	                       | 4224                                  | 53537                        | 0.08            |
|100 kb	     |1031                         | 1883                                  | 26768                        | 0.07            |
Table: Domain data and class imbalance summaries across resolutions for GM12878.

@@

A total of 78 cell type specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 1). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). We constructed three different predictor types for each functional genomic annotation (Figure 1B). Counts of overlaps (**Overlap Counts**, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (**Overlap Percent**, OP) were defined as the proportion of a bin overlapped by an annotation region. **Distance** was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering steps were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (OP), and indirect (distance) relationships between TAD boundaries and genomic annotations.

@@

## Building an ensemble framework for TAD boundary prediction

To account for the heavily imbalanced classes reported above, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. **Random over-sampling** (ROS) was defined as sampling with replacement from TAD boundary regions. **Random under-sampling** (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered **Synthetic minority over-sampling technique** (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

To identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction, we created an ensemble framework to systematically compare performances of different model inputs (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using a particular predictor type *P = {OC, OP, Distance}*, and re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on chromosome-specific datasets using 5-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 3 \times 4 \times 21 = 2520$) models. <!--Additional performance metrics are reported in the Supplementary materials.-->

![**A diagram of the ensemble framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the rowise (genomic bins) and columnwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)). Step 2 is to split **A** into a training and testing set according to a 7:3 ratio. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to implement the random forest algorithm and perform 5-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data for unbiased model evaluation.](Figures/ensemble_framework.png)
